\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-17 k-planar.tex
Cleaned-up 1-planar proof

The remainder of the paper is organized as follows: In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{1-planar} we prove \thmref{1-planar}.  
In \secref{g-k-planar} we prove \thmref{g-k-planar}.

@@

  Finally, let $M$ and $M^+$ denote the subgraph of $G$ containing the edges and vertices of $Y$, respectively $Y^+$, and the edges and vertices of $F$.  The graph $M^+$ has some number of bounded faces, all contained in the interior of $F$. Some of the bounded faces of $M^+$ are kite faces. Let $F_1,\ldots,F_m$ be the non-kite bounded faces of $M^+$.

  We claim that, for each $i\in\{1,\ldots,m\}$, the boundary of $F_i$ is a cycle in $G'$.  Otherwise, some edge $vw$ contributes to the boundary of $F_i$ but is crossed by an edge $xy\in E(G)$. Then, $vw\not\in E(F)$ since no edge of $F$ is crossed by any edge of $G$. Therefore $vw\in E(\overline{Y}^+)$ so $xy\in E(Y^+)$. But then the only faces of $M^+$ incident to $vw$ are kite faces.  In particular $vw$ can not be incident the non-kite face $F_i$.

@@

  The remainder of the proof finishes in the same way as the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $V(Y^+)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.

  The planarity of $H$ comes from two properties:
  \begin{enumerate}
    \item $G/\mathcal{P}$ and $G'/\mathcal{P}$ are isomorphic.  This follows immediately from the fact that, for each edge $xy\in E(G-G')$, $x,y\in S_a$ for some $a\in V(H)$.  Indeed, $S_a=Y^+_a$ where $\overline{Y}_a$ contains the edge $vw\in E(G')$ that crosses $xy$.

    \item $G[\overline{Y}^+-F]$ is connected. To see why this is so, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y}^+$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
  \end{enumerate}
  Since $G'$ is planar, this implies that $H=G'/\mathcal{P}$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}
\end{proof}

Finally, we can complete the proof of \thmref{1-planar}.

\begin{proof}[Proof of \thmref{1-planar}]
  Using \lemref{induction}, the proof of \thmref{1-planar} is now straightforward. Given a 1-plane graph $G$, add edges to make it edge maximal so that it has an outer face $F=v_1v_2v_3$. Next, add a vertex $r$ adjacent to $v_1$, $v_2$, and $v_3$ to obtain an edge-maximal 1-plane graph $\overline{G}$ with one vertex $r$ of degree 3 on its outer face.

@@

  Use the layering $\mathcal{L}=\langle V_0',\ldots,V_h'\rangle$ where $V_i'=V_{2i}\cup V_{2i+1}$ for each $i\in\{0,\ldots,h\}$. That this is a layering of $G$ follows from \eqref{spar-distance}.  Since $|V_j\cap S_x|\le 15$ for every integer $i\ge 0$, $|V_i'\cap S_x|\le 30$ for every integer $i\ge 0$ and $x\in V(H)$.
\end{proof}

@@

@@

% \section{Consequences}
% \seclabel{consequences}

>>>>

The remainder of the paper is organized as follows: In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{1-planar} we prove \thmref{1-planar}.  
In \secref{g-k-planar} we prove \thmref{g-k-planar}. In \secref{consequences} we point out some consequences of these results for some of the other graph parameters mentioned above.

@@

  Let $M$ and $M^+$ denote the subgraph of $G$ containing the edges and vertices of $\overline{Y}$, respectively $\overline{Y}^+$, and the edges and vertices of $F$.  The graph $M^+$ has some number of bounded faces, all contained in the interior of $F$. Some of the bounded faces of $M^+$ are kite faces. Let $F_1,\ldots,F_m$ be the non-kite bounded faces of $M^+$.

  We claim that, for each $i\in\{1,\ldots,m\}$, the boundary of $F_i$ is a cycle in $G'$ that contains no spars. Otherwise, some edge $vw$ contributes to the boundary of $F_i$ but is crossed by an edge $xy\in E(G)$. Then, $vw\not\in E(F)$ since no edge of $F$ is crossed by any edge of $G$. Therefore $vw\in E(\overline{Y}^+)$ so $xy\in E(Y^+)$. But then the only faces of $M^+$ incident to $vw$ are kite faces.  In particular $vw$ can not be incident the non-kite face $F_i$.

@@

  The remainder of the proof finishes in the same way as the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $S=V(Y^+)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}, except that now each node $x$ has a child for each face $F_i$ of $M^+_x$ that contains a vertex of $G$ in its interior.

  The planarity of $H$ comes from two properties:
  \begin{enumerate}
    \item $G/\mathcal{P}$ and $G^+/\mathcal{P^+}$ are isomorphic, where $G^+$ is the triangulation obtained by adding dummy vertices at each crossing in $G$ and $\mathcal{P}^+$ is the partition we obtain by adding a dummy vertex $z$ to $\overline{Y}^+$ if $\overline{Y}^+$ contains an edge $vw$ that contains $z$ in its interior.

    \item $G^+[\overline{Y}^+-F]$ is connected. To see why this is so, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y}^+$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
  \end{enumerate}
  Since $G^+$ is planar, the second point implies that $H=G^+/\mathcal{P}$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.} The first point then ensures that $G/\mathcal{P}$ is also planar.
\end{proof}

Using \lemref{induction}, the proof of \thmref{1-planar} is now straightforward.

\begin{proof}[Proof of \thmref{1-planar}]
Given a 1-plane graph $G$, add edges to make it edge maximal so that it has an outer face $F=v_1v_2v_3$. Next, add a vertex $r$ adjacent to $v_1$, $v_2$, and $v_3$ to obtain an edge-maximal 1-plane graph $\overline{G}$ with one vertex $r$ of degree 3 on its outer face.

@@

  Use the layering $\mathcal{L}=\langle V_0',\ldots,V_h'\rangle$ where $V_i'=V_{2i}\cup V_{2i+1}$ for each $i\in\{0,\ldots,h\}$. That this is a layering of $G$ follows from \eqref{spar-distance}.  Since $|V_j\cap S_x|\le 15$ for every integer $i\ge 0$, $|V_i'\cap S_x|\le 30$ for every integer $i\ge 0$ and every $x\in V(H)$.
\end{proof}

@@

Next we consider a generalization of our results to $(g,k)$-plane graphs.

@@

\section{Consequences}
\seclabel{consequences}

\begin{enumerate}
  \item $p$-centered colourings;
  \item non-repetitive chromatic number;
  \item connection with queue layouts;
  \item connection with low-treewidth/treedepth colourings
\end{enumerate}
